---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# RPracticePackage

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/heidisteadman/RPracticePackage)](https://github.com/heidisteadman/RPracticePackage/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/heidisteadman/RPracticePackage)](https://github.com/heidisteadman/RPracticePackage/pulls)
<!-- badges: end -->

The goal of `RPracticePackage` is to help the developers gain practice creating Bioconductor packages with experimental data.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `RPracticePackage` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("RPracticePackage")
```

This code installs BiocManager if it is not already installed. Running the if statement is required to install the package as it is dependent on BiocManager to load. 

Download the development version from [GitHub](https://github.com/heidisteadman/RPracticePackage) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("heidisteadman/RPracticePackage")
```

## Data Info

The data used in this package was accessed through Open Science Framework (OSF). This database allowed us to use data that was already neat and ready to be analyzed. However, the original data came from Gene Expression Omnibus (GEO). The name of every SummarizedExperiment object is the data set name in GEO. Learn more about the data at the following links:

 - GSE41197: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE41197 
 - GSE10797: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE10797 
 - GSE59772: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE59772 

The package itself includes a metadata as required by Bioconductor. This file is a table that lists the name of each object and more information about where it came from and how it was accessed. See inst/extdata/metadata.csv. When the package is available in ExperimentHub, using the query() function will show available objects. 

```{r load_metadata}
#library(ExperimentHub)
#eh = ExperimentHub()
#query(eh, RPracticePackage)

# this code throws errors because the package is not currently available in ExperimentHub

```


## Examples

The following is the code required to access the objects within the package, as well as additional packages we used to analyze the data. The SummarizedExperiment package is required for both building the package and accessing the data. 
```{r load_packages}
library(RPracticePackage)
library(SummarizedExperiment)
library(tidyverse)
```

The SummarizedExperiment object for each data set includes 3 matrices. The first, which we load as expression_data, is a matrix with the samples as columns and genes as rows. The Ensembl gene ID is used as the row names. The data in the actual matrix is the expression level of the gene for the sample. <br>
<br>
The second matrix we load as sample_metadata. This is information about the samples themselves, such as demographics for the patient the sample came from. The samples are rows and the information is in the columns. <br>
<br>
The third matrix we load as feature_data. This is the information about the genes that were tested for each sample. This includes information like the chromosome the gene is located on, the Entrez gene ID, and the gene name. The genes are the rows and the information about the samples is in the columns.

```{r example, eval = requireNamespace('RPracticePackage')}
data('GSE41197')
expression_data = assay(GSE41197)
sample_metadata = colData(GSE41197)
feature_data = rowData(GSE41197)
```

This is a simple example of using a built-in function to get a general idea of the data. The summary() function includes data points for each sample, like the minimum or maximum expression levels across all the genes. Generally, this isn't particularly relevant, but does give an idea of how many samples are in the matrix.

```{r experiment_summary}
data('GSE41197')
gene_expression_values = assay(GSE41197)
summary(gene_expression_values)
```

In the following example, we first access the matrix of gene expression values. To make it easier to analyze, we convert it into a tibble and select the first 10 rows. We then used ggplot() to create a bar plot showing the gene expression levels for one sample across 10 genes.

```{r plot_example}
data('GSE41197')
gene_expression_values = assay(GSE41197)

exp_tib = as_tibble(gene_expression_values, rownames='Ensembl_Gene_ID')[1:10,] %>%
    ggplot(aes(x=Ensembl_Gene_ID, y=GSM1010328)) +
    geom_col() +
    labs(x='Ensembl Gene ID',y='Sample ID GSM1010328') +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

exp_tib
```


Here, we first accessed the matrices for 3 data sets included in the RPracticePackage. 

```{r plot_example_2}
data('GSE10797')
data('GSE41197')
data('GSE59772')

GSE10197 = assay(GSE10797)
GSE41197 = assay(GSE41197)
GSE59772 = assay(GSE59772)
```

Next, we converted each to a tibble and selected one row, representing one gene that we knew each data set had in common. 

```{r plot_example_2_part_2}
GSE10197_gene = GSE10197 %>%
    as_tibble(rownames='Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID == "ENSG00000171428")
GSE41197_gene = GSE41197 %>%
    as_tibble(rownames='Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID == "ENSG00000171428") 
GSE59772_gene = GSE59772 %>%
    as_tibble(rownames='Ensembl_Gene_ID') %>%
    filter(Ensembl_Gene_ID == "ENSG00000171428")
```

We then joined the 3 tibbles into one tibble with the full_join() function. Then, we rotated the tibble to make the selected gene be the column and the samples be the rows. We now have one tibble with one column (the gene) and many rows (the samples). <br>
<br>
Finally, we used ggplot() to visualize the expression levels for the gene across the three data sets.

```{r plot_example_2_part_3}
combined_samples = full_join(GSE10197_gene, GSE41197_gene)
combined_samples = full_join(combined_samples, GSE59772_gene) %>%
    pivot_longer(cols=-Ensembl_Gene_ID,names_to='Sample_ID',values_to='Expression_Level') %>%
    ggplot(aes(x=Sample_ID,y=Expression_Level)) +
    geom_col() +
    #theme_bw() +
    labs(x='Sample ID',y='Expression level of ENSG00000171428') +
    coord_flip() +
    scale_fill_brewer(palette = "Set3") +
    theme_bw() +
    theme(axis.text.y = element_text(size = 6))
combined_samples
```

## Citation

Below is the citation output from using `citation('RPracticePackage')` in R. Please
run this yourself to check for any updates on how to cite __RPracticePackage__.

```{r 'citation', eval = requireNamespace('RPracticePackage')}
print(citation('RPracticePackage'), bibtex = TRUE)
```

Please note that the `RPracticePackage` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://heidisteadman.github.io/RPracticePackage) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.

## Code of Conduct
  Please note that the RPracticePackage project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.


