---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# RPracticePackage

<!-- badges: start -->
[![GitHub issues](https://img.shields.io/github/issues/heidisteadman/RPracticePackage)](https://github.com/heidisteadman/RPracticePackage/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/heidisteadman/RPracticePackage)](https://github.com/heidisteadman/RPracticePackage/pulls)
<!-- badges: end -->

The goal of `RPracticePackage` is to help the developers gain practice creating Bioconductor packages with experimental data.

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `RPracticePackage` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("RPracticePackage")
```

And the development version from [GitHub](https://github.com/heidisteadman/RPracticePackage) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("heidisteadman/RPracticePackage")
```
## Example

Load the data into the SummarizedExperiment object:

```{r example, eval = requireNamespace('RPracticePackage')}
library("RPracticePackage")
library(SummarizedExperiment)
data('se')
expression_data = assay(se)
sample_metadata = colData(se)
feature_data = rowData(se)
```

Example of a summary of the expression data:

```{r experiment_summary}
library("RPracticePackage")
library(SummarizedExperiment)
data('GSE41197')
expressions = assay(GSE41197)
summary(expressions)
```

Example of the first 10 gene expression levels of the first sample read into a plot:

```{r plot_example}
library("RPracticePackage")
library(tidyverse)
library(SummarizedExperiment)

data('GSE41197')
expressions = assay(GSE41197)

exp_tib = as_tibble(expressions, rownames='Ensembl_Gene_ID')[1:10,] %>%
    ggplot(aes(x=Ensembl_Gene_ID, y=GSM1010328)) +
    geom_col() +
    labs(x='Ensembl Gene ID',y='Sample ID GSM1010328') +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

exp_tib
```


## Citation

Below is the citation output from using `citation('RPracticePackage')` in R. Please
run this yourself to check for any updates on how to cite __RPracticePackage__.

```{r 'citation', eval = requireNamespace('RPracticePackage')}
print(citation('RPracticePackage'), bibtex = TRUE)
```

Please note that the `RPracticePackage` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `RPracticePackage` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://heidisteadman.github.io/RPracticePackage) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.

## Code of Conduct
  Please note that the RPracticePackage project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.


